---
title: "Aggregated Summary"
format:
  html:
    fig-width: 8
    embed-resources: false
vignette: >
  %\VignetteIndexEntry{Aggregated Summary}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`reactableShiny` provides four Shiny modules that layer interactive inputs
on top of `{reactable}` tables. Each module follows the same two-function
pattern — a `*_ui()` function and a `*_server()` function — and returns a
reactive that your app can consume downstream.

Custom visual overrides belong in `www/custom.css`, which Shiny serves
automatically.

---

## 1. `flexible_table` — user-built editable rows

**When to use it.** The user constructs a table from scratch by choosing
from dropdowns, typing free text, or entering dates. Rows can be added,
deleted, and reset. Use this when the dataset does not exist yet and the
user is the author.

`flexible_table_ui()` returns a **named list** (`table`, `add_row_button`,
`reset_button`) so each element can be placed anywhere in your layout
independently.

```{r flexible-table, eval=FALSE}
library(shiny)
library(bslib)
library(reactableShiny)

variety_map <- list(
  apple  = c("Fuji" = "fuji", "Gala" = "gala"),
  banana = c("Cavendish" = "cavendish"),
  pear   = c("Bosc" = "bosc", "Bartlett" = "bartlett")
)

col_specs <- list(
  list(
    name    = "fruit",
    label   = "Fruit",
    type    = "select",
    choices = c("Select..." = "", "Apple" = "apple",
                "Banana" = "banana", "Pear" = "pear")
  ),
  list(
    name                 = "variety",
    label                = "Variety",
    type                 = "select",
    choices              = c("Select..." = ""),
    depends_on           = "fruit",
    get_filtered_choices = function(val) {
      c("Select..." = "", variety_map[[val]] %||% character(0))
    }
  ),
  list(
    name        = "notes",
    label       = "Notes",
    type        = "text",
    placeholder = "Optional...",
    width       = 220
  )
)

ui <- page_sidebar(
  sidebar = sidebar(
    flexible_table_ui("picker")$add_row_button,
    br(),
    flexible_table_ui("picker")$reset_button
  ),
  flexible_table_ui("picker")$table,
  card(
    card_header("Selected (complete rows only)"),
    verbatimTextOutput("flexible_result")
  )
)

server <- function(input, output, session) {
  selected <- flexible_table_server(
    id                 = "picker",
    col_specs          = col_specs,
    duplicates_allowed = FALSE
  )
  output$flexible_result <- renderPrint(selected())
}

shinyApp(ui, server)
```

**Column spec quick reference**

| Field | Required? | Purpose |
|---|---|---|
| `name` | Yes | Data frame column name & Shiny input ID prefix |
| `label` | Yes | Header text |
| `type` | No* | `"select"` (default), `"text"`, `"date"` |
| `choices` | select | Named character vector or grouped named list |
| `default` | No | Pre-filled value for new rows (`""` if omitted) |
| `width` | No | Column width in pixels |
| `depends_on` | No | Name of a parent column whose value filters this one |
| `get_filtered_choices` | conditional | `function(parent_value)` returning a choices vector |

*Omitting `type` defaults to `"select"` for backward compatibility.

The server returns a reactive **data frame of complete, non-blank rows**.
Internal helper columns (`id`, `delete`) are stripped automatically.
Pass `initial_data = reactive(df)` to pre-populate on startup.

---

## 2. `annotator_super_simple` — lightweight row selection

**When to use it.** You have an existing data frame and want the user to
flag a subset of rows with a single click. No per-cell inputs — the only
output is a character vector of selected row IDs.

`annotator_super_simple_ui()` returns a named list (`table`,
`reset_button`).

```{r annotator-super-simple, eval=FALSE}
library(shiny)
library(bslib)
library(reactableShiny)

queue <- reactive({
  data.frame(
    id      = 1:6,
    patient = c("Alice", "Bob", "Carol", "Dave", "Eve", "Frank"),
    status  = c("pending", "urgent", "pending", "done", "urgent", "pending"),
    age     = c(34, 52, 29, 61, 45, 38),
    stringsAsFactors = FALSE
  )
})

ui <- page_sidebar(
  sidebar = sidebar(
    annotator_super_simple_ui("review")$reset_button
  ),
  annotator_super_simple_ui("review")$table,
  card(
    card_header("Selected IDs"),
    verbatimTextOutput("simple_result")
  )
)

server <- function(input, output, session) {
  selected_ids <- annotator_super_simple_server(
    id                = "review",
    source_data       = queue,
    row_id            = "id",
    display_cols      = c("patient", "status", "age"),
    selection         = "multiple",
    reactable_options = list(searchable = TRUE, striped = TRUE)
  )
  output$simple_result <- renderPrint(selected_ids())
}

shinyApp(ui, server)
```

**Key behaviour notes**

- Selections persist across reactive updates to `source_data`. If a row
  disappears (e.g. due to a filter) and reappears, it stays selected.
- The return value contains only IDs *currently visible* in `source_data`.
- Pass additional `reactable()` arguments via
  `reactable_options = list(...)`. Do not include `data`, `selection`,
  `onClick`, `rowStyle`, `columns`, `defaultSelected`, `elementId`, or
  `theme` — these are owned by the module.

---

## 3. `annotator_table` — per-row structured annotation

**When to use it.** You have an existing data frame and want the user to
fill in structured metadata for each row: dropdowns, checkboxes, text
fields, numbers, or radio buttons.

`annotator_table_ui()` returns a single `reactableOutput()` element.

The example below also demonstrates **gating**: the `notes` field is
greyed out and non-interactive until the `approved` checkbox in the same
row is checked.

```{r annotator-table, eval=FALSE}
library(shiny)
library(bslib)
library(reactableShiny)

source_data <- reactive({
  data.frame(
    car = c("Mazda RX4", "Hornet 4 Drive", "Valiant", "Merc 240D", "Fiat 128"),
    mpg = c(21.0, 21.4, 18.1, 24.4, 32.4),
    stringsAsFactors = FALSE
  )
})

col_specs <- list(
  list(name = "car", type = "display", label = "Car"),
  list(name = "mpg", type = "display", label = "MPG"),
  list(
    name    = "category",
    type    = "select",
    label   = "Category",
    choices = c("Economy" = "economy", "Sport" = "sport", "Luxury" = "luxury")
  ),
  list(name = "approved", type = "checkbox", label = "Approved?"),
  list(
    name        = "notes",
    type        = "text",
    label       = "Notes",
    placeholder = "Reason for approval...",
    gates       = "approved"
  )
)

ui <- page_fluid(
  h4("Car Annotation"),
  annotator_table_ui("cars"),
  br(),
  card(
    card_header("Touched rows only"),
    verbatimTextOutput("annotator_result")
  )
)

server <- function(input, output, session) {
  result <- annotator_table_server(
    id          = "cars",
    source_data = source_data,
    row_id      = "car",
    col_specs   = col_specs
  )
  output$annotator_result <- renderPrint(result())
}

shinyApp(ui, server)
```

**Column type reference**

| `type` | Widget | Extra required fields |
|---|---|---|
| `"display"` | Read-only text | — |
| `"clickable_display"` | Clickable cell that toggles a checkbox | `toggles` |
| `"select"` | Dropdown | `choices` |
| `"checkbox"` | Checkbox | — |
| `"text_checkbox"` | Clickable label + checkbox | `display_col` |
| `"text"` | Free-text input | — |
| `"number"` | Numeric input | — |
| `"radio"` | Radio button group | `choices` |

All types accept `label`, `width`, `col_def_options`, and `gates` (the
name of a checkbox column that controls this input's interactivity).

The server returns a reactive **data frame of the `row_id` column plus
all input annotation columns**, filtered to rows where at least one input
has been touched. Display columns are excluded. Pass
`initial_values = reactive(df)` to pre-seed annotations.

---

## 4. `milestone_tracker` — two-panel select-then-annotate

**When to use it.** Your data is grouped (age bands, project phases,
departments) and you want the user to first pick items from a grouped
table on the left, then annotate only those selected items on the right.
This module composes `annotator_table` internally.

`milestone_tracker_ui()` returns a complete two-column `bslib` layout —
no individual element placement needed.

```{r milestone-tracker, eval=FALSE}
library(shiny)
library(bslib)
library(reactableShiny)

milestones <- reactive({
  data.frame(
    id         = 1:8,
    age        = c("6 months",  "6 months",  "12 months", "12 months",
                   "18 months", "18 months", "24 months", "24 months"),
    age_months = c(6, 6, 12, 12, 18, 18, 24, 24),
    milestone  = c(
      "Sits with support", "Babbles",
      "Pulls to stand",    "Says mama",
      "Walks alone",       "Points to objects",
      "Runs steadily",     "Uses 2-word phrases"
    ),
    domain     = c("Motor", "Language", "Motor", "Language",
                   "Motor", "Language", "Motor", "Language"),
    stringsAsFactors = FALSE
  )
})

annotation_specs <- list(
  list(name = "milestone", type = "display", label = "Milestone"),
  list(name = "domain",    type = "display", label = "Domain"),
  list(
    name    = "status",
    type    = "select",
    label   = "Status",
    choices = c(
      "Achieved" = "achieved",
      "Emerging" = "emerging",
      "Not yet"  = "not_yet"
    )
  ),
  list(name = "notes", type = "text", label = "Notes")
)

ui <- page_fluid(
  milestone_tracker_ui("tracker"),
  br(),
  card(
    card_header("Touched annotations"),
    verbatimTextOutput("tracker_result")
  )
)

server <- function(input, output, session) {
  result <- milestone_tracker_server(
    id                   = "tracker",
    data                 = milestones,
    row_id               = "id",
    group_col            = "age",
    group_sort_col       = "age_months",
    item_col             = "milestone",
    badge_col            = "domain",
    badge_colors         = list(
      Motor    = list(bg = "#EAF3DE", text = "#27500A"),
      Language = list(bg = "#DEE9F3", text = "#0A2750")
    ),
    annotation_col_specs = annotation_specs,
    card1_title          = "Select Milestones",
    card2_title          = "Record Progress",
    count_label          = "milestones"
  )
  output$tracker_result <- renderPrint(result())
}

shinyApp(ui, server)
```

**Key behaviour notes**

- Clicking a group header row in Card 1 selects *all* rows in that group.
- Card 2 only appears once at least one row is selected.
- Annotation state is preserved when a row is deselected and re-selected.
- `group_sort_col` controls group sort order (hidden from display).
  When `NULL`, groups sort alphabetically by `group_col`.

---

## Quick reference

| Module | Source data | Returns |
|---|---|---|
| `flexible_table` | User-built | `data.frame` of complete, non-blank rows |
| `annotator_super_simple` | Provided | `character` vector of selected row IDs |
| `annotator_table` | Provided | `data.frame` of touched annotation rows |
| `milestone_tracker` | Provided | `data.frame` of touched annotation rows (selected items only) |

All server functions accept `reactable_theme` (a `reactableTheme` object)
and `reactable_options` (a named list of extra `reactable()` arguments) to
customise rendering without modifying module internals.
