---
title: "Annotator Table"
format:
  html:
    fig-width: 8
    column: screen-inset
vignette: >
  %\VignetteIndexEntry{Annotator Table}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

The annotator table is a Shiny module that layers interactive input columns on
top of a read-only data frame. It is designed for human-in-the-loop workflows
where you want reviewers to tag, categorise, or score every row in a dataset —
without being able to edit the underlying data.

## How it works

You provide a source data frame and a list of **column specifications**
(`col_specs`). Each spec declares whether the column should be a read-only
display column or an interactive input — `select`, `checkbox`, `text`,
`number`, or `radio`. The module renders everything in a single
[reactable](https://glin.github.io/reactable/) table and returns a reactive
data frame of annotations for rows that have been touched by the user.

A few things worth knowing:

- **Annotations survive filtering.** If `source_data` is a reactive expression
  that can change (e.g. driven by a search box), annotations for rows that
  temporarily disappear are preserved and restored when those rows return.
- **Only touched rows are returned.** The module's return value excludes rows
  where every input is still at its default, so you don't have to filter out
  blanks yourself.
- **Gating is built in.** Any input column can declare `gates = "my_checkbox"`
  to have its interactivity controlled by a checkbox in the same row. When the
  checkbox is unchecked the gated inputs grey out instantly — no Shiny
  round-trip needed.

## Basic usage

Attach the UI placeholder with `annotator_table_ui()` and start the server
logic with `annotator_table_server()`. The server function returns a reactive
that you can observe or pass downstream.

```{r basic-example, eval=FALSE}
library(shiny)
library(bslib)
library(reactableShiny)

col_specs <- list(
  list(name = "car",      type = "display",  label = "Car"),
  list(name = "mpg",      type = "display",  label = "MPG"),
  list(name = "cyl",      type = "display",  label = "Cylinders"),
  list(name = "category", type = "select",   label = "Category",
       choices = c("Economy" = "economy", "Sport" = "sport", "Truck" = "truck")),
  list(name = "approved", type = "checkbox", label = "Approved?"),
  list(name = "notes",    type = "text",     label = "Notes",
       placeholder = "Optional comment...")
)

ui <- page_fluid(
  h4("Review cars"),
  annotator_table_ui("cars"),
  hr(),
  h4("Annotations so far"),
  tableOutput("result")
)

server <- function(input, output, session) {
  cars_df <- tibble::rownames_to_column(mtcars, var = "car")
  source  <- reactive(cars_df)

  annotations <- annotator_table_server(
    id          = "cars",
    source_data = source,
    row_id      = "car",
    col_specs   = col_specs
  )

  output$result <- renderTable(annotations())
}

shinyApp(ui, server)
```

Running the app produces a table with the `car`, `mpg`, and `cyl` columns
shown as plain text, and a **Category** dropdown, **Approved?** checkbox, and
**Notes** field in each row. `output$result` updates in real time to show only
the rows where at least one input has been filled in.

## Column types

| Type | Returns | Key fields |
|---|---|---|
| `display` | — (read-only) | `name` must exist in `source_data` |
| `select` | `character` | `choices` (named character vector) |
| `checkbox` | `logical` | — |
| `text` | `character` | `placeholder` (optional) |
| `number` | `numeric` | `min`, `max` (optional) |
| `radio` | `character` | `choices` (named character vector) |

Every column type also accepts an optional `label` (shown as the column
header; defaults to `name`) and `width` in pixels.

## Gating example

Use `gates` to make one checkbox control whether other inputs in the same row
are interactive. This is useful when a "flag this row" checkbox should unlock
additional detail fields.

```{r gating-example, eval=FALSE}
col_specs <- list(
  list(name = "car",    type = "display",  label = "Car"),
  list(name = "flag",   type = "checkbox", label = "Flag for review?"),
  list(name = "reason", type = "select",   label = "Reason",
       choices = c("Price" = "price", "Safety" = "safety"),
       gates   = "flag"),
  list(name = "notes",  type = "text",     label = "Notes",
       placeholder = "Details...",
       gates   = "flag")
)
```

Here, **Reason** and **Notes** are greyed out until **Flag for review?** is
checked. Unchecking it again restores the previous values — they are never
cleared.

## Interactive demo

The code below is the full self-contained app used in the live demo. Copy it
into a new R session and run `shinyApp(ui, server)` to try it interactively.
It uses a representative set of developmental milestone rows: check **Met?**
to mark a milestone as achieved, and add a short note — the annotation panel
updates in real time to show only rows you have touched.

```{r milestone-demo, eval=FALSE}
library(shiny)
library(bslib)
library(reactableShiny)

milestones <- data.frame(
  id        = 1:10,
  age       = c("6 months",  "6 months",  "12 months", "12 months",
                "18 months", "18 months", "24 months", "24 months",
                "36 months", "36 months"),
  category  = c("Motor", "Language", "Motor", "Language",
                "Motor", "Language", "Motor", "Language",
                "Motor", "Language"),
  milestone = c(
    "Sits with support",          "Babbles consonants",
    "Pulls to stand",             "Says mama/dada with meaning",
    "Walks alone",                "Points to objects of interest",
    "Runs without falling",       "Uses 2-word phrases",
    "Jumps with both feet",       "Speaks in 3-word sentences"
  ),
  stringsAsFactors = FALSE
)

col_specs <- list(
  list(name = "id",        type = "display",  label = "ID",        width = 50),
  list(name = "age",       type = "display",  label = "Age",       width = 110),
  list(name = "category",  type = "display",  label = "Category",  width = 120),
  list(name = "milestone", type = "display",  label = "Milestone", width = 340),
  list(name = "met",       type = "checkbox", label = "Met?",      width = 60),
  list(
    name        = "notes",
    type        = "text",
    label       = "Notes",
    placeholder = "Short note...",
    width       = 200
  )
)

ui <- page_fluid(
  theme = bs_theme(version = 5),
  card(
    card_header("Developmental Milestones"),
    card_body(annotator_table_ui("milestones"))
  ),
  card(
    card_header("Annotations so far"),
    card_body(tableOutput("annotations_out"))
  )
)

server <- function(input, output, session) {
  source_data <- reactive(milestones)

  annotations <- annotator_table_server(
    id          = "milestones",
    source_data = source_data,
    row_id      = "id",
    col_specs   = col_specs
  )

  output$annotations_out <- renderTable({
    ann <- annotations()
    if (nrow(ann) == 0) {
      return(data.frame(Message = "No milestones annotated yet."))
    }
    dplyr::left_join(ann, milestones, by = "id") |>
      dplyr::select(id, age, category, milestone, met, notes)
  })
}

shinyApp(ui, server)
```
