---
title: "Annotator Table"
output: rmarkdown::html_vignette
runtime: shiny
vignette: >
  %\VignetteIndexEntry{Annotator Table}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

The annotator table is a Shiny module that layers interactive input columns on top of a read-only data frame. It is designed for human-in-the-loop workflows where you want reviewers to tag, categorise, or score every row in a dataset — without being able to edit the underlying data.

## How it works

You provide a source data frame and a list of **column specifications** (`col_specs`). Each spec declares whether the column should be a read-only display column or an interactive input — `select`, `checkbox`, `text`, `number`, or `radio`. The module renders everything in a single [reactable](https://glin.github.io/reactable/) table and returns a reactive data frame of annotations for rows that have been touched by the user.

A few things worth knowing:

- **Annotations survive filtering.** If `source_data` is a reactive expression that can change (e.g. driven by a search box), annotations for rows that temporarily disappear are preserved and restored when those rows return.
- **Only touched rows are returned.** The module's return value excludes rows where every input is still at its default, so you don't have to filter out blanks yourself.
- **Gating is built in.** Any input column can declare `gates = "my_checkbox"` to have its interactivity controlled by a checkbox in the same row. When the checkbox is unchecked the gated inputs grey out instantly — no Shiny round-trip needed.

## Basic usage

Attach the UI placeholder with `annotator_table_ui()` and start the server logic with `annotator_table_server()`. The server function returns a reactive that you can observe or pass downstream.




```{r example}
library(shiny)
library(bslib)
library(reactableShiny)

col_specs <- list(
  list(name = "car",      type = "display",  label = "Car"),
  list(name = "mpg",      type = "display",  label = "MPG"),
  list(name = "cyl",      type = "display",  label = "Cylinders"),
  list(name = "category", type = "select",   label = "Category",
       choices = c("Economy" = "economy", "Sport" = "sport", "Truck" = "truck")),
  list(name = "approved", type = "checkbox", label = "Approved?"),
  list(name = "notes",    type = "text",     label = "Notes",
       placeholder = "Optional comment...")
)

ui <- page_fluid(
  h4("Review cars"),
  annotator_table_ui("cars"),
  hr(),
  h4("Annotations so far"),
  tableOutput("result")
)

server <- function(input, output, session) {
  # Add a row-identifier column — mtcars uses row names
  cars_df <- tibble::rownames_to_column(mtcars, var = "car")

  source <- reactive(cars_df)

  annotations <- annotator_table_server(
    id          = "cars",
    source_data = source,
    row_id      = "car",
    col_specs   = col_specs
  )

  output$result <- renderTable(annotations())
}

shinyApp(ui, server)
```

Running the app produces a table with the `car`, `mpg`, and `cyl` columns shown as plain text, and a **Category** dropdown, **Approved?** checkbox, and **Notes** field in each row. `output$result` updates in real time to show only the rows where at least one input has been filled in.

## Column types

| Type | Returns | Key fields |
|---|---|---|
| `display` | — (read-only) | `name` must exist in `source_data` |
| `select` | `character` | `choices` (named character vector) |
| `checkbox` | `logical` | — |
| `text` | `character` | `placeholder` (optional) |
| `number` | `numeric` | `min`, `max` (optional) |
| `radio` | `character` | `choices` (named character vector) |

Every column type also accepts an optional `label` (shown as the column header; defaults to `name`) and `width` in pixels.

## Gating example

Use `gates` to make one checkbox control whether other inputs in the same row are interactive. This is useful when a "flag this row" checkbox should unlock additional detail fields.

```{r specs}
col_specs <- list(
  list(name = "car",      type = "display",  label = "Car"),
  list(name = "flag",     type = "checkbox", label = "Flag for review?"),
  list(name = "reason",   type = "select",   label = "Reason",
       choices = c("Price" = "price", "Safety" = "safety"),
       gates   = "flag"),
  list(name = "notes",    type = "text",     label = "Notes",
       placeholder = "Details...",
       gates   = "flag")
)
```

Here, **Reason** and **Notes** are greyed out until **Flag for review?** is checked. Unchecking it again restores the previous values — they are never cleared.



```{r module, echo=FALSE}
library(dplyr)
library(purrr)
library(reactable)
library(reactable.extras)
library(reactableShiny)
```



```{css echo=FALSE}
/* -------------------------------------------------------
   Annotator table demo styles
   ------------------------------------------------------- */

.rt-td,
.rt-th {
  vertical-align: middle !important;
  padding: 6px 10px !important;
}

.card {
  margin-bottom: 1.25rem;
  border: 1px solid #dee2e6;
}

.card-header {
  font-weight: 600;
  background-color: #f8f9fa;
  border-bottom: 1px solid #dee2e6;
}

.form-control[type="text"] {
  font-size: 0.875rem;
  padding: 2px 6px;
  height: auto;
}

.form-control::placeholder {
  color: #adb5bd;
}

/* Break the app container out of the document's content column */
.breakout {
  width: 100vw;
  position: relative;
  left: 50%;
  transform: translateX(-50%);
  padding: 0 1rem;
}
```


```{r load, echo=FALSE}

milestones <- arrow::read_parquet(
  system.file("datasets/df_developmental_milestones.parquet", package = "reactableShiny")
)

```


```{r app, echo=FALSE}
library(reactable)
library(reactable.extras)
library(reactableShiny)
library(dplyr)



# chunk: col_specs
col_specs <- list(
  list(name = "id",        type = "display",  label = "ID",        width = 60),
  list(name = "age",       type = "display",  label = "Age",       width = 110),
  list(name = "category",  type = "display",  label = "Category",  width = 180),
  list(name = "milestone", type = "display",  label = "Milestone", width = 420),
  list(name = "met",       type = "checkbox", label = "Met?",      width = 60),
  list(name = "notes",     type = "text",     label = "Notes",
       placeholder = "Short note...",          width = 200)
)

# chunk: shiny app
ui <- bslib::page_fluid(
  theme = bslib::bs_theme(version = 5),

  htmltools::div(
    class = "breakout",

    bslib::card(
      bslib::card_header("Developmental Milestones"),
      bslib::card_body(
        annotator_table_ui("milestones")
      )
    ),

    bslib::card(
      bslib::card_header("Annotations so far"),
      bslib::card_body(
        tableOutput("annotations_out")
      )
    )
  )
)

server <- function(input, output, session) {
  source_data <- shiny::reactive(milestones)

  annotations <- annotator_table_server(
    id          = "milestones",
    source_data = source_data,
    row_id      = "id",
    col_specs   = col_specs
  )

  output$annotations_out <- shiny::renderTable({
    ann <- annotations()

    # Guard: nothing touched yet → friendly message rather than empty table
    if (nrow(ann) == 0) {
      return(data.frame(Message = "No milestones annotated yet."))
    }

    # Join back age/category/milestone so the reviewer sees context,
    # not just a bare ID and checkbox value
    purrr::reduce(
      list(ann, milestones),
      ~ dplyr::left_join(.x, .y, by = "id")
    ) |>
      dplyr::select(id, age, category, milestone, met, notes)
  })
}


```





```{r, out.width='100vw', out.height="800px", out.extra='style="margin-left: calc(50% - 50vw); max-width: none;"'}
shinyApp(ui, server)
```


