---
title: "Using the flexible_table module"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the flexible_table module}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

The `flexible_table` r shiny module provides an easy way to create dynamic data entry.
Each row contains one dropdown per column. Rows can be added, deleted, and
reset. The module returns a reactive data frame of the current selections,
which updates live as the user interacts with the table.


## Simple Example: People Who Live in Your Home

Let's say you have a form that asks users who lives in their home.

A short list of possibilities of who could live in a users home is:


*  mother
*  father
*  grandmother
*  grandfather
*  sibling
*  sibling (male)
*  sibling (female)
*  step-mother
*  step-father
*  partner
*  spouse

```{r}
home_occupants <- c("mother", "father", "grandmother", "grandfather", "sibling", "sibling (male)", "sibling (female)", "step-mother", "step-father", "partner", "spouse")
```


### To specify 


Column specifications

Every column in the table is described by a **column spec** — a named list
with at minimum a `name`, `label`, and `choices`:

```{r}
col_specs <- list(
  list(
    name    = "home_occupants",
    label   = "Home Occupants",
    choices = setNames(
      c(1:length(home_occupants)),
      home_occupants
    )
  )
)
```

The first entry in each `choices` vector is a blank placeholder — a value of
`""` with a label of `""`. This is the default empty state for a new row.

> **Note:** R does not allow `"" = ""` as inline named-vector syntax. Use
> `setNames()` to assign an empty string as a name.

### Wiring up the UI

`flexible_table_ui()` returns a **named list** rather than a single UI
element. This lets you place the table and its buttons independently in your
layout:

```{r}
ui <- bslib::page_fluid(
  bslib::layout_columns(
    col_widths = c(2, 10),
    div(
      flexible_table_ui("my_table")$add_row_button,
      br(),
      flexible_table_ui("my_table")$reset_button
    ),
    flexible_table_ui("my_table")$table
  )
)
```

If you want all three elements together in one place, `do.call(tagList, ...)`
works as a shortcut:

```{r}
ui <- bslib::page_fluid(
  do.call(tagList, flexible_table_ui("my_table"))
)
```

### Wiring up the server

`flexible_table_server()` returns a **reactive data frame** of the current
table contents. Rows where every cell is blank are excluded. Assign the return
value to a variable to use it downstream:

```{r}
server <- function(input, output, session) {

  selected <- flexible_table_server(
    id        = "my_table",
    col_specs = col_specs
  )

  # selected() is a reactive data frame — use it like any other reactive
  observe({
    print(selected())
  })

}
```

### Complete minimal example

```{r}
library(shiny)
library(bslib)
library(reactable.shiny)

home_occupants <- c("mother", "father", "grandmother", "grandfather", "sibling", "sibling (male)", "sibling (female)", "step-mother", "step-father", "partner", "spouse")

col_specs <- list(
  list(
    name    = "home_occupant_id",
    label   = "Home Occupants",
    choices = setNames(
      c(1:length(home_occupants)),
      home_occupants
    )
  )
)

ui <- page_fluid(
  do.call(tagList, flexible_table_ui("example")),
  verbatimTextOutput("result")
)

server <- function(input, output, session) {
  selected <- flexible_table_server("example", col_specs)

  output$result <- renderPrint(selected())
}

shinyApp(ui, server)
```


In the above example, you can see that a user can add as many 'siblings' or 'mothers' or 'fathers' as they like.
If, however, you want to disallow duplicates, you can do that as well.

```{r}
library(shiny)
library(bslib)
library(reactable.shiny)

home_occupants <- c("mother", "father", "grandmother", "grandfather", "sibling", "sibling (male)", "sibling (female)", "step-mother", "step-father", "partner", "spouse")

col_specs <- list(
  list(
    name    = "home_occupant_id",
    label   = "Home Occupants",
    choices = setNames(
      c(1:length(home_occupants)),
      home_occupants
    )
  )
)

ui <- page_fluid(
  do.call(tagList, flexible_table_ui("example")),
  verbatimTextOutput("result")
)

server <- function(input, output, session) {
  selected <- flexible_table_server("example", col_specs, duplicates_allowed = FALSE)

  output$result <- renderPrint(selected())
}

shinyApp(ui, server)
```



---

## Pre-populating with `initial_data`

Pass a reactive data frame to `initial_data` to pre-fill the table. This is
useful when editing an existing record — you load the saved values into the
table, the user makes changes, and you read the updated values back out.

```{r}
server <- function(input, output, session) {

  # A reactive that returns the data to load — could come from a database,
  # a file, or another reactive in your app


  selected <- flexible_table_server(
    id           = "my_table",
    col_specs    = col_specs,
    initial_data = existing_data
  )

}
```


```{r}
library(shiny)
library(bslib)
library(reactable.shiny)

home_occupants <- c("mother", "father", "grandmother", "grandfather", "sibling", "sibling (male)", "sibling (female)", "step-mother", "step-father", "partner", "spouse")



col_specs <- list(
  list(
    name    = "home_occupant_id",
    label   = "Home Occupants",
    choices = setNames(
      c(1:length(home_occupants)),
      home_occupants
    )
  )
)

ui <- page_fluid(
  do.call(tagList, flexible_table_ui("example")),
  verbatimTextOutput("result")
)

server <- function(input, output, session) {
    existing_data <- reactive({
    data.frame(
      home_occupant_id = c(2,1)
    )
  })
    
  selected <- flexible_table_server(
    "example", 
    col_specs, 
    initial_data = existing_data,
    duplicates_allowed = FALSE)

  output$result <- renderPrint(selected())
}

shinyApp(ui, server)
```


Column names in `initial_data` must match the `name` fields in `col_specs`.

`initial_data` is **reactive** — if the reactive expression fires again (for
example, because the user switches to a different record), the table is
replaced with the new data automatically.

### Resetting to a blank table

Passing a data frame with zero rows triggers a reset to a single blank row.
This is useful when switching from edit mode to add mode:

```{r}
existing_data <- reactive({
  if (input$mode == "add") {
    data.frame()          # zero rows — signals a reset
  } else {
    load_record(input$record_id)
  }
})
```

---

## Options

### `duplicates_allowed`

Set `duplicates_allowed = FALSE` to silently remove duplicate rows from the
returned data frame. Two rows are considered duplicates if every column value
is identical:

```{r}
selected <- flexible_table_server(
  id                 = "my_table",
  col_specs          = col_specs,
  duplicates_allowed = FALSE
)
```

Note that duplicates are only removed from the *returned* reactive — the user
can still see and interact with them in the table itself.

### `complete_rows_only`

By default the module returns rows where **at least one** value has been
chosen. Completely empty rows — such as the initial blank row before the user
touches anything — are excluded.

If your downstream code requires every column to be filled in, set
`complete_rows_only = TRUE`:

```{r}
selected <- flexible_table_server(
  id                 = "my_table",
  col_specs          = col_specs,
  complete_rows_only = TRUE
)
```

The table below summarises how each setting affects which rows are returned:

| `complete_rows_only` | Rows returned |
|---|---|
| `FALSE` (default) | Any row with at least one non-blank value |
| `TRUE` | Only rows where every column is filled in |

### Combining both options

The two options compose independently:

```{r}
selected <- flexible_table_server(
  id                 = "my_table",
  col_specs          = col_specs,
  duplicates_allowed = FALSE,
  complete_rows_only = TRUE
)
```

This returns only fully complete rows, with any duplicates among them removed.

---

## Applying a custom theme

The table uses `theme_bare` by default — a minimal transparent theme that
inherits styling from the surrounding page. To apply a different
[`reactable::reactableTheme()`](https://glin.github.io/reactable/reference/reactableTheme.html),
pass it to `reactable_theme`:

```{r}
my_theme <- reactable::reactableTheme(
  backgroundColor = "#f8f9fa",
  borderColor     = "#dee2e6"
)

selected <- flexible_table_server(
  id              = "my_table",
  col_specs       = col_specs,
  reactable_theme = my_theme
)
```
